Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-09-02 vertical-thermal-review.Rmd
small changes

$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).

@@

$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).

@@

Canopy trees account for the majority of forest ecosystem water and carbon cycling.
While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, where the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. 
Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed ~93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. 
In terms of C cycling, it has been estimated that the canopy strata contributes ≥64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007]. 
Large trees also dominate woody aboveground net primary productivity (*ANPP~woody~*) and mortality (*M~woody~*), with trees ≥10cm DBH usually contributing > 85% of *ANPP~woody~* and *M~woody~* across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].

![**Figure `r fig_ecosystem`. Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamicsinpress.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)

>>>>

$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer); TrS = tropical savanna.

@@

$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN = boreal needleleaf (conifer).

@@

Canopy trees account for the majority of forest ecosystem water and carbon cycling.
While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, where the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. 
Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed ~93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. 
In terms of C cycling, it has been estimated that the canopy strata contributes ≥64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007]. 
Large trees also dominate woody aboveground net primary productivity (*ANPP~woody~*) and mortality (*M~woody~*), with trees ≥10cm DBH usually contributing > 85% of *ANPP~woody~* and *M~woody~* across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiotDistributionBiomassDynamics2022].

![**Figure `r fig_ecosystem`. Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamics2022.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
